---
title: "Bioc.gff: GFF3 File Format Support"
author: "Michael Lawrence"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioc.gff: GFF3 File Format Support}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: true
package: Bioc.gff
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    cache = TRUE,
    comment = "#>"
)
library(BiocStyle)
```

# Bioc.gff

## Introduction

The `r Biocpkg("Bioc.gff")` package provides support for the General Feature
Format (GFF) file format, which is widely used for representing genomic features
and annotations. This package allows users to read, write, and manipulate GFF
versions 1, 2 (GTF), and 3 in R, making it easier to work with genomic data.

Note that much of the code in the package is adapted from the
`r Biocpkg("rtracklayer")` Bioconductor package. The `Bioc.gff` package is
designed to be a lightweight alternative, focusing specifically on GFF file
handling without the additional features of `rtracklayer`.

# Installation

You can install the `Bioc.gff` package from Bioconductor using the following:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Bioc.gff")
```

# Loading packages

```{r load}
library(Bioc.gff)
library(GenomicRanges)
```

# Usage

The `Bioc.gff` package provides functions to read and write GFF files, as well
as to manipulate GFF data structures. The following sections provide some
examples of how to use the package.

## Reading GFF Files

You can read GFF files using the `readGFF` function. This function supports
GFF versions 1, 2, and 3. The following example demonstrates how to read a GFF
file:

```{r find-gff}
gff_file <- system.file("extdata", "genes.gff3", package = "Bioc.gff")
```

### Using the `import` function

The `import` function deduces that the input string is a GFF file type and calls
the appropriate method for reading the GFF file.

```{r read-gff}
import(gff_file)
```

Here the output object is a `GRanges` class, which contains genomic coordinates
and associated metadata. The `GRanges` object indicates the ranges with the
`seqnames`, `ranges`, and `strand`,  columns. The associated metadata is
stored in the `mcols` slot, which (in this example) includes attributes like
`source`, `type`, `phase`, `ID`, `Name`, `geneName`, `Alias` and `genome`.

#### Selectively importing ranges

You can also selectively import ranges from the GFF file using the `which`
argument. This allows you to filter the data based on specific criteria, such as
chromosome or strand. For example, to import only the ranges on chromosome 1:

```{r read-gff-which}
which <- GRanges("chr10:90000-93000")
import(gff_file, which = which, genome = "hg19")
```

Note that you can indicate the genome build using the `genome` argument, which
is useful for ensuring that the imported data is compatible with other genomic
data you may be working with.

### Using the `readGFF` function

As an alternative, you can use the `readGFF` function directly, which is
more explicit about the GFF version:

```{r read-gff-version}
readGFF(gff_file, version = "3")
```

This function returns a `DataFrame` object, which contains the same information
as the `GRanges` object but in a tabular format. Note that one can use the
`makeGRangesFromDataFrame` function to convert the `DataFrame` returned by
`readGFF` into a `GRanges` object, which is often more convenient for downstream
analysis:

```{r read-gff-granges}
readGFF(gff_file, version = "3") |>
    makeGRangesFromDataFrame(
        keep.extra.columns = TRUE
    )
```

### Reading remote GFF files

You can also read GFF files from remote URLs. The `import` function can handle
URLs directly, allowing you to read GFF files hosted on remote servers. 
On the Bioconductor Build System (BBS), we use the `BiocFileCache` package
to cache remote files, which avoids downloading the file every time the code
gets run.

```{r read-gff-remote}
library(BiocFileCache)
bfc <- BiocFileCache::BiocFileCache()
url <- "https://www.mirbase.org/download/hsa.gff3"
bquery <- bfcquery(bfc, url, "rname", exact = TRUE)
if (!nrow(bquery))
    bfcadd(x = bfc, rname = url, rtype = "web", download = TRUE)
gff_remote <-
    bfcrpath(bfc, rnames = url, exact = TRUE, download = FALSE, rtype = "web")
readGFF(gff_remote)
```

Otherwise, you can read the remote GFF file directly using the `readGFF`
function:

```{r read-gff-remote-direct,eval=FALSE}
readGFF(url, version = "3")
```

## Conversion to GFF

```{r as-gff}
library(GenomicFeatures)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
exonsBy(txdb, by = "tx") |>
    asGFF()
```

## GFF to TxDb

You can convert a `GFF` object to a `TxDb` object using the `makeTxDbFromGFF`
in the `r Biocpkg("txdbmaker")` package. This is useful for creating a
transcript database from GFF annotations, which can then be used for various
genomic analyses.

```{r make-txdb-from-gff}
library(txdbmaker)
txdb <- makeTxDbFromGFF(
    file = gff_remote,
    format = "gff3",
    dataSource = "https://www.mirbase.org/download/hsa.gff3",
    organism = "Homo sapiens",
    taxonomyId = 9606
)
genome <- grepv("genome-build-id", readLines(gff_remote)) |>
    strsplit("# genome-build-id:\\s+") |>
    unlist() |>
    tail(1L)
genome(txdb) <- genome
txdb
```

# SessionInfo

```{r session-info}
sessionInfo()
```
